>

### 2019-10-01 manuscript/manuscript.Rmd
Rewrite literature review - initial
This edit restructures the presentation of the utility-based accessibility measure to focus more clearly on extensions and similarities with the gravity model. Need to revise references accordingly.

---
title: >
  Modeling the impacts of park access on health outcomes: a choice-based 
  accessibility approach
author:
  - name: Gregory Macfarlane
    email: gregmacfarlane@byu.edu
    affiliation: Brigham Young University
    corresponding: gregmacfarlane@byu.edu
  - name: Nico Boyd
    email:
    affiliation: Fehr and Peers 
  - name: John E. Taylor
    email: jetaylor@gatech.edu
    affiliation: Georgia Institute of Technology
  - name: Kari Watkins
    email: kariwatkins@gatech.edu
    affiliation: Georgia Institute of Technology
address:
  - code: Brigham Young University
    address: Civil and Environmental Engineering Department, Brigham Young University 430 Engineering Building, Provo, UT, 84602
  - code: Fehr and Peers
    address: Los Angeles, California
  - code: Georgia Institute of Technology
    address: School of Civil and Environmental Engineering, Georgia Institute of Technology, 755 Atlantic Drive, Atlanta, GA, 30332
abstract: >
  Recent research has underscored the potential for public green spaces to
  influence individual and societal health outcomes, but empirical measurements
  of such influences have yielded mixed results to date, with particular
  disagreement surrounding how access to parks ought to be defined while
  controlling alternate explanations. In this paper we present a comprehensive
  measure of park accessibility drawn from random utility choice theory, which
  avoids arbitrary assertions of proximity while incorporating potentially
  numerous amenities and attributes of both the parks and the population. We
  apply this choice-based accessibility measure to correlate Census tract-level
  obesity and physical activity rate estimates from the Centers for Disease
  Control and Prevention's 500 Cities project with tract-level American
  Community Survey socioeconomic data in New York City, paired with geographic
  parks data from New York City. Controlling for the
  socioeconomic variables and spatially correlated error terms, we show a
  positive and significant relationship between park access and physical
  activity rates, and a clear and significant negative relationship between park
  access and obesity rates. In doing so, this research contributes a more
  comprehensive modeling approach for measuring the impact of park access on
  health, and may improve our understanding of the role  parks and access to
  them can serve in furthering public health objectives.
author_summary: >
  Greg Macfarlane is an assistant professor of Civil Engineering at Brigham
  Young University in Provo, Utah.

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
Over the last few years a new social model of health has evolved, treating health
``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences.''[@Duhl1999, p.7]
In this new paradigm, resources provided via civil infrastructure --- in
particular parks and other public green spaces --- play a crucial role in
promoting and sustaining public health [@Bedimo-Rung2005; @Coutts2008].
After all, it does little to lecture individuals on the importance of exercise
if their community is built in such a way that such exercise is expensive,
unenjoyable, or unsafe.

In spite of the potential for green spaces to influence public health outcomes 
and numerous studies attempting to empirically estimate the strength of these
outcomes, evidence to date has been mixed. A major challenge researchers have
faced is a proliferation of techniques for measuring access to green spaces, of
which many are plausible but lack robust theoretical basis. In this
paper, we present a holistic and flexible measurement for park accessibility
based in activity location choice theory. This choice-based accessibility measure
compares the continuous distance to all parks in the region, weighted against
the sizes of the each park and other assorted amenities.

@@

We also demonstrate how the choice-based accessibility metric can potentially be
expanded to account for other measures of a park's attractiveness including its
observed social media activity and the presence of various park amenities.
The paper concludes with a discussion of opportunities for future research.

@@

Numerous previous studies have found relationships between characteristics of
the built environment and aspects of physical, mental, social, and economic
health. For the purposes of this study, it is clear that parks and other
vegetated areas support outdoor physical activity in ways distinct from other
urban environments [@Giles-Corti2005; @Wells2007].  In spite of these generally
positive findings, the evidence of a specific link between green space and
physical activity is somewhat mixed.

<!-- Rewrite this section: 
  - Cross-metropolitan studies are somewhat clear?
  - Within-metropolitan studies are limited
-->

@Wolf2008 reports that people self-identify as more active and attain a
higher quality of life when a greater portion of their environment is greenspace.
@Larson2016 similarly used self-reported scores on the Gallup-Healthways
Wellbeing Index to evaluate the relationship between different areas of
wellbeing, including physical health, and park quantity, quality, and
accessibility in 44 U.S. cities. While the authors found positive associations
between wellbeing and park quality and accessibility, these relationships were
not statistically significant. A study by @West2012 used park data from
the Trust for Public Land’s 2010 City Park Facts and public health data from the
Behavioral Risk Factor Surveillance System (BRFSS) to examine relationships
between the density of parkland, parkland per capita, and levels of physical
activity and obesity for 67 metropolitan statistical areas in the U.S. The study
found a significant, positive association between park density and physical
activity and a significant, negative association between park density and
obesity. Conversely, @Richardson2012 examined the relationship between
urban greenspace and selected mortality rates, and though the authors did not
find a statistically significant relationship between the quantity of urban
greenspace and mortality caused by lung cancer, diabetes, heart disease or car
accidents, they did find that all-cause mortality was oddly, *higher* in
greener cities. In a meta-analysis of 20 peer reviewed journal articles
exploring the relationship between parks and objectively measured physical
activity, @Bancroft2015 found that five studies reported a significant
positive association between the two, six studies produced mixed results, and
nine studies found no association at all.

While several of the aforementioned studies reported results consistent with the
hypothesis that park access and use confer health benefits stemming from
increased physical activity and attendant decreases in obesity, the large
majority of extant research has been conducted at the level of the city or
metropolitan statistical area. Metropolitan-level analyses do not capture the
within-city variation in accessibility to parks that exist in many cities. Some
cities with large amounts of total greenspace may nevertheless unequally
distribute this space throughout the city, leading to areas with poor access.
Conversely, a city with a smaller overall proportion of greenspace may give all
of its citizens better access to sufficient greenspace to meet their needs or
wants for physical activity. Considering access at the neighborhood level within
a single city may also eliminate some regional or cultural fixed effects
affecting metropolitan-level analyses.

In a study of neighborhoods in New York City, @Stark2014 used
hierarchical linear models to explore relationships between the proportion of
zip codes that was park space, the cleanliness of parks, and the body mass index
of adult residents. The authors found that a higher proportion of park space in
a neighborhood was associated with a lower BMI for its residents, and lower park
cleanliness scores were associated with a higher BMI, suggesting that increasing
the supply of clean, well-maintained parks could positively impact adult BMI.

## Choice-based Accessibility

Consider that an individual is choosing a park for a recreation activity.
If we apply random utility choice theory, specifically the multinomial
logit model [@McFadden1974], the expected probability of individual
$i$ choosing park $j$ from the set of all regional parks $J$ is
\begin{equation}\label{eq:mnl}
P(j | V_{ij}) = \frac{\exp(V_{ij})}{\sum_{p \in J}\exp(V_{ip})}
\end{equation}
where parks are differentiated from each other by their relative measurable
utilities $V_{ij}$. In principle, $V$ may include any measurable attributes of either
the choice maker or the park. In this study we use a linear formulation of
\begin{equation}\label{eq:v_sd}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d
\end{equation}
incorporating the size of the park in acres and the distance of the park from
the census tract $i$ in miles. The coefficients $\vect{\lambda}$ can be
estimated from household surveys, though in the absence of a survey we
may assert reasonable values.

A key theoretical outcome of random utility choice models is that the
consumer surplus to individual $i$, or the total weighted value of all
alternatives in individual $i$'s choice set, can
be obtained as the logarithm of the sum in the denominator of the choice
probability equation [@Small1981]:
\begin{equation}\label{eq:cs}
C_i = \ln\left({\sum_{p \in J}\exp(V_{ip})}\right)
\end{equation}
There are several advantages to such a log-sum defined metric relative to
buffer-based accessibility metrics more commonly found in the literature. First,
all individuals are defined as having some access to all parks, rather than an
arbitrary cutoff asserted by the researcher.
This allows for the fact that some people are more or less sensitive to
distances, and that distance is a continuous, non-binary phenomenon.
It defies reason to assume a person living 1.1 miles
from a park has functionally different accessibility than someone living at 0.9 miles.
Second, the random utility formulation
allows the researcher to include any attribute of the park as part of its utility
specification; in this case, we consider the size of the park weighted against its
distance as an element of relative accessibility. This suggests that
not all parks are equal, and that a large park such as Central Park in Manhattan
may provide health and activity benefits over a much larger area than a smaller
community square.

Note that this formulation is an extension of gravity-based accessibility
statistics, in the same way that the gravity-based trip distribution model
is a specific case of
a destination-choice model [@Daly1982]. But the extension is meaningful and
allows us to consider more attributes of the park or the individual beyond size
and impedance. For example, we can aggregate the number of tweets emanating from
a park as a measure of the park's attractiveness beyond its size. The
utility for the choice maker from Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d + tweets_j\lambda_t
\end{equation}

In spite of its flexibility and basis in choice theory, accessibility measures
based in log-sums have not received much application in the accessibility
literature compared with distance-based or even gravity-based measures.
@Zhang2011 present a metric with a similar functional form, though the
authors arrive at it from a motivation other than destination choice theory.
Log-sums are commonly used, however, in alternatives analyses of transit
infrastructure improvements [@DeJong2007]. A reason for this is likely that
a regional travel demand model is available to the analysts, thus making
calibrated and multimodal logsums readily available [@Geurs2010].

>>>>

---
title: >
  Modeling the impacts of park access on health outcomes: a utility-based 
  accessibility approach
author:
  - name: Gregory Macfarlane
    email: gregmacfarlane@byu.edu
    affiliation: Brigham Young University
    corresponding: gregmacfarlane@byu.edu
  - name: Nico Boyd
    email:
    affiliation: Fehr and Peers 
  - name: John E. Taylor
    email: jetaylor@gatech.edu
    affiliation: Georgia Institute of Technology
  - name: Kari Watkins
    email: kariwatkins@gatech.edu
    affiliation: Georgia Institute of Technology
address:
  - code: Brigham Young University
    address: Civil and Environmental Engineering Department, Brigham Young University 430 Engineering Building, Provo, UT, 84602
  - code: Fehr and Peers
    address: Los Angeles, California
  - code: Georgia Institute of Technology
    address: School of Civil and Environmental Engineering, Georgia Institute of Technology, 755 Atlantic Drive, Atlanta, GA, 30332
abstract: >
  Recent research has underscored the potential for public green spaces to
  influence individual and societal health outcomes, but empirical measurements
  of such influences have yielded mixed results to date, with particular
  disagreement surrounding how access to parks ought to be defined while
  controlling alternate explanations. In this paper we present a comprehensive
  measure of park accessibility drawn from random utility choice theory, which
  avoids arbitrary assertions of proximity while incorporating potentially
  numerous amenities and attributes of both the parks and the population. We
  apply this choice-based accessibility measure to correlate Census tract-level
  obesity and physical activity rate estimates from the Centers for Disease
  Control and Prevention's 500 Cities project with tract-level American
  Community Survey socioeconomic data in New York City, paired with geographic
  parks data from New York City. Controlling for the
  socioeconomic variables and spatially correlated error terms, we show a
  positive and significant relationship between park access and physical
  activity rates, and a clear and significant negative relationship between park
  access and obesity rates. In doing so, this research contributes a more
  comprehensive modeling approach for measuring the impact of park access on
  health, and may improve our understanding of the role  parks and access to
  them can serve in furthering public health objectives.
author_summary: >
  Greg Macfarlane is an assistant professor of Civil Engineering at Brigham
  Young University in Provo, Utah.

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
Over the last several years a new social model of health has evolved, treating health
``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences.''[@Duhl1999, p.7]
In this new paradigm, resources provided via civil infrastructure --- in
particular parks and other public green spaces --- play a crucial role in
promoting and sustaining public health [@Bedimo-Rung2005; Wells2007; @Coutts2008].
After all, it does little to lecture individuals on the importance of exercise
if their community is built in such a way that such exercise is expensive,
unenjoyable, or unsafe.

In spite of the potential for green spaces to influence public health outcomes 
and numerous studies attempting to empirically estimate the strength of these
outcomes, evidence to date has been mixed. A major challenge researchers have
faced is a proliferation of techniques for measuring access to green spaces, of
which many are plausible but lack robust theoretical basis. In this
paper, we present a holistic and flexible measurement for park accessibility
based in activity location choice theory. This utility-based accessibility measure
compares the continuous distance to all parks in the region, weighted against
the sizes of each park and its other assorted amenities.

@@

We also demonstrate how the utility-based accessibility metric can potentially be
expanded to account for other measures of a park's attractiveness including its
observed social media activity and the presence of various park amenities.
The paper concludes with a discussion of opportunities for future research.

@@

The large majority of extant research on the topic of park access and public health
outcomes has been involved comparing metropolitan regions against each other. For example,
a study by @West2012 used park data from
the Trust for Public Land’s 2010 City Park Facts and public health data from the
Behavioral Risk Factor Surveillance System (BRFSS) to examine the relationship
between the density of park land and physical activity and obesity rates for 67
metropolitan statistical areas in the U.S. The findings in this case conformed to 
expectation; that is, the study found a significant positive association
between park density and physical activity and a significant negative
association between park density and obesity.
@Larson2016 similarly used self-reported scores on the Gallup-Healthways
Wellbeing Index to evaluate the relationship between physical health and park
quantity, quality, and accessibility in 44 U.S. cities. While the authors found
positive associations between wellbeing and park quality and accessibility,
these relationships were not statistically significant.
Conversely, @Richardson2012 examined the relationship between total urban
greenspace and mortality rates for selected maladies; though the authors did not
find a statistically significant relationship between the quantity of urban
greenspace and mortality caused by lung cancer, diabetes, heart disease or car
accidents, they did find that all-cause mortality was oddly, *higher* in greener
cities. Finally, in a meta-analysis of 20 peer reviewed journal articles exploring the
relationship between parks and objectively measured physical activity,
@Bancroft2015 found that five studies reported a significant positive
association between the two, six studies produced mixed results, and nine
studies found no association at all.

Metropolitan-level analyses such as those described above do not capture the
within-city variation in accessibility to parks that exist in many cities. Some
cities with large amounts of total greenspace may nevertheless unequally
distribute this space throughout the city, leading to areas with poor access.
Conversely, a city with a smaller overall proportion of greenspace may give all
of its citizens better access to sufficient greenspace to meet their needs or
wants for physical activity. Considering access at the neighborhood level within
a single city may also eliminate some regional or cultural fixed effects
affecting metropolitan-level analyses. In a study of neighborhoods in New York
City, @Stark2014 found that a higher proportion of park space in a neighborhood
was associated with a lower BMI for its residents, and lower park cleanliness
scores were associated with a higher BMI, suggesting that increasing the supply
of clean, well-maintained parks could positively impact adult BMI.

## Park Accessibility Measures

Loosely defined, the *accessibility* of an arbitrary place describes the ease
with which people can accomplish activities there. Accessibility is an abstract
concept with tempting quantifiability, perhaps explaining the proliferation of
calculation mechanisms. @Dong2006 present a helpful mathematical heirarchy of
common mechanisms, which we follow here. Consider a person residing at
point $i$ in a city with parks $j \in 1 \ldots J$. An analyst might consider
point $i$ as "having access" to a park if the distance to the park $d_ij$  is less than some threshold
$D$,

\begin{equation}\ref{eq:isochrone}
A_i = \begin{cases} 
      1 & d_{ij}\leq D\\
      0 & d_{ij} > D
   \end{cases}
\end{equation}

Variations of this isochrone-based framework are easily derived and relatively common. 
@ParkScore is a sophisticated application of this approach where $d_{ij}$ is a
carefully calculated 10-minute network-bound walk, with a demographic analysis of
areas within and without this threshold. In other applications $D$ might be more variably defined,
such as "Does a park exist within the census tract?". The indicator function could also be
modified into the total number of parks or percent of park space within radius $D$ of point $i$.

A somewhat more holistic approach is the so-called "gravity" accessibility
statistic. In this case the accessibility of a point is the denominator of the 
gravity formulation of a trip distribution model,

\begin{equation}\ref{eq:gravity}
A_i = \sum_{j = 1}^J S_j f(d_{ij})
\end{equation}

where $S_j$ is the "size" of each park literally (i.e. acres) or abstractly (i.e.
trip attractions) and $f(d_ij)$ is a monotonically decreasing cost function,
usually a negative exponential. @Dong2006 note that the isochrone framework is a
special case of the gravity model where $f(d_{ij})$ is a binary function.

A third accessibility mechanism is a utility-based specification, termed as such
for being derived from random utility choice theory. Consider that an individual
living at point $i$ is choosing a park for a recreation activity. If we apply the multinomial logit
model [@McFadden1974], the expected consumer surplus enjoyed by this individual 
can be shown to be 
\begin{equation}\label{eq:utility}
A_i = \ln\left({\sum_{j = 1}^J\exp(V_{ij})}\right) + C
\end{equation}
where parks are differentiated from each other by their relative measurable
*utilities* $V_{ij}$. $C$ is an unknown constant, but the difference in consumer
surplus between two points $i$ and $k$ can be quantitatively compared as 
$A_i - A_k$. In principle, $V$ may include any measurable attributes of either
the choice maker or the park, and is typically represented as a linear-in-parameters
function of destination attributes $X_j$ and the travel cost $d_{ij}$,
\begin{equation}\label{eq:utility}
V_{ij} = X_j\beta + \beta_d * d_{ij}
\end{equation}
the census tract $i$ in miles. The coefficients $\vect{\beta}$ can be
are frequently estimated from household surveys, though in the absence of a survey we
may assert reasonable values.

Note that the gravity formulation is itself a special case of the utility-based
specification where $\exp(X_j\beta + \beta_d d_{ij}) = X_j f(d_{ij})$ [@Daly1982].
The distinction between gravity and utility-based specification is meaningful, 
however. Primarily, it becomes possible to construct accessibility statistics based
on revealed behavioral preferences rather than calibrated or asserted values.

Both gravity and utility-based specifications hold several advantages relative to
isochrone-based accessibility metrics more commonly found in the literature. First,
all individuals are defined as having some access to all parks, rather than an
arbitrary cutoff asserted by the researcher.
This allows for the fact that some people are more or less sensitive to
distances, and that distance is a continuous, non-binary phenomenon.
It defies reason to assume a person living 1.1 miles
from a park has functionally different accessibility than someone living at 0.9 miles.
Second, the random utility formulation
allows the researcher to include any attribute of the park as part of its utility
specification; in this case, we consider the size of the park weighted against its
distance as an element of relative accessibility. This suggests that
not all parks are equal, and that a large park such as Central Park in Manhattan
may provide health and activity benefits over a much larger area than a smaller
community square.

In spite of its flexibility and basis in choice theory, accessibility measures
based in choice maker utility have not received much application in the accessibility
literature compared with distance-based or even gravity-based measures.
@Vale2016 also present a descriptive classification of
active accessibility mechanisms, though these authors curiously dismiss
utility-based metrics for incorporating randomness. The accessibility
formula presented in \ref{eq:utility} is the expectation of a random utility process,
and is not in and of itself random any more than the gravity model.
@Zhang2011 present a metric with a similar functional form, though the
authors arrive at it from a motivation other than destination choice theory.
Log-sums are commonly used, however, in alternatives analyses of transit
infrastructure improvements [@DeJong2007]. A reason for this is likely that
a regional travel demand model is available to the analysts, thus making
calibrated and multimodal logsums readily available [@Geurs2010].
